---
title: "Investigating Factors Assoicated with Alzheimer's Diagnosis"
author: "Ben daSilva Curiel 35207323 \n, Dylan Forsythe 47473699 \n, Yiran Wang 55134043"
date: "`r Sys.Date()`"
output: pdf_document
---

## Introduction
 
 -  The Alzheimer's Disease:
 
Alzheimer's disease is a progressive brain disease and the most common cause and type of dementia. Dementia causes a decline in memory and cognitive abilities, and in severe cases can even affect basic functions of daily life. The disease also shortens the lifespan of patients and places long-term physical and mental burdens on patients and their families. More than half of all dementia patients in the world suffer from Alzheimer's disease, and the disease is currently incurable, the interventions only slowing its progression. *(3)*

One well-known gene that influences Alzheimer’s risk is the apolipoprotein E (APOE) gene. The APOE gene is involved in making a protein that helps carry cholesterol and other types of fat in the bloodstream. Problems in this process are thought to contribute to the development of Alzheimer’s. APOE comes in several forms, called alleles (e.g., $\epsilon2, \epsilon3$).
*(1) (directly from National Institute on Aging)*

As the population ages, the incidence of Alzheimer's disease continues to rise. It is estimated that by 2030, there will be 78 million people suffering from dementia worldwide, and by 2050, there will be 139 million patients. This will put huge pressure on the medical system and social resources. Therefore, early identification and judgment of disease risk is not only crucial for patients and their families to better deal with the disease, but also contributes to a more rational allocation of medical resources in the society. Given its widespread impact, prevention and early identification of the disease are particularly important.

This article selects a data set on Alzheimer’s Diagnosis’ (Ankit 2025) on Kaggle.com on Feb. 6, 2025 to analyze. This data set was created by Ankush Panday, and originally authored by Ankit. This data was released under the MIT license. According to the kaggle webpage, Ankush Panday utilized web scraping techniques to collect the data from two health reports in 2023 and will be updated on an annual basis. This means that the data may come from public health databases, research papers, or online resources of government or medical institutions, rather than from surveys or experiments. There is no missing data or duplicate values found.

The data set contains 74,283 records and 25 variables, and the intended purpose is to study the factors associated with an Alzheimer’s diagnosis. It includes patients from 20 distinct countries and includes demographic information, health status, lifestyle, socioeconomic background, and genetic factors.

Overall, this data set contains a wide variety of clinical and demographic features that are related to Alzheimer’s diagnosis worldwide, making it suitable for exploring potential risk factors for Alzheimer’s disease and supports data analysis and modeling simulation.

Our motivation for this research is preventative, to help identify possible lifestyle and genetic indicators of a positive Alzheimer's diagnosis and understand the desparing reality of the disease.

```{r Librarysetup, include=FALSE}
library(mgcv)
library(dplyr)
library(ggplot2)
library(splines)
library(car)
library(MASS)
library(parallel)
library(knitr)
library(tidyr)
library(boot)
library(DAAG)
library(caret)
library(knitr)
```

```{r DataImport,include=FALSE}
setwd("/Users/Bendasilva/Desktop/410_Project")
alzData<-read.csv("alzheimers_prediction_Dataset.csv",header = TRUE,sep=",")
```

```{r ColNameChange,include=FALSE}
## The conversion
# Step 1: Rename columns to match the simplified names
colnames(alzData)[colnames(alzData) == "Family.History.of.Alzheimer.s"] <- "FamilyHistory"
colnames(alzData)[colnames(alzData) == "Genetic.Risk.Factor..APOE.ε4.allele."] <- "GeneticRisk"
colnames(alzData)[colnames(alzData) == "Alzheimer.s.Diagnosis"] <- "AlzheimerDiagnosis"
colnames(alzData)[colnames(alzData) == "Physical.Activity.Level"] <- "PhysicalActivityLevel"
colnames(alzData)[colnames(alzData) == "Smoking.Status"] <- "SmokingStatus"
colnames(alzData)[colnames(alzData) == "Cholesterol.Level"] <- "CholesterolLevel"
colnames(alzData)[colnames(alzData) == "Cognitive.Test.Score"] <- "CognitiveTestScore"
colnames(alzData)[colnames(alzData) == "Sleep.Quality"] <- "SleepQuality"
colnames(alzData)[colnames(alzData) == "Air.Pollution.Exposure"] <- "AirPollutionExposure"
colnames(alzData)[colnames(alzData) == "Marital.Status"] <- "MaritalStatus"
colnames(alzData)[colnames(alzData) == "Social.Engagement.Level"] <- "SocialEngagementLevel"
colnames(alzData)[colnames(alzData) == "Stress.Levels"] <- "StressLevels"
colnames(alzData)[colnames(alzData) == "Education.Level"] <- "EducationLevel"
colnames(alzData)[colnames(alzData) == "Alcohol.Consumption"] <- "AlcoholConsumption"
colnames(alzData)[colnames(alzData) == "Depression.Level"] <- "DepressionLevel"
colnames(alzData)[colnames(alzData) == "Dietary.Habits"] <- "DietaryHabits"
colnames(alzData)[colnames(alzData) == "Employment.Status"] <- "EmploymentStatus"
colnames(alzData)[colnames(alzData) == "Income.Level"] <- "IncomeLevel"
colnames(alzData)[colnames(alzData) == "Urban.vs.Rural.Living"] <- "UrbanvsRuralLiving"

# Step 2: Update factor_vars list to use the new column names
factor_vars <- c("Country", "Gender", "PhysicalActivityLevel", "SmokingStatus",
                 "AlcoholConsumption", "Diabetes", "Hypertension", "CholesterolLevel", 
                 "FamilyHistory", "DepressionLevel", "SleepQuality", "DietaryHabits", 
                 "AirPollutionExposure", "EmploymentStatus", "MaritalStatus", 
                 "GeneticRisk", "SocialEngagementLevel", "IncomeLevel", "StressLevels", 
                 "UrbanvsRuralLiving", "AlzheimerDiagnosis")

# Step 3: Check if all factor variables exist in the data
missing_columns <- setdiff(factor_vars, colnames(alzData))
if(length(missing_columns) > 0) {
  stop(paste("Missing columns:", paste(missing_columns, collapse = ", ")))
}

# Step 4: Convert the specified columns to factors (after renaming)
alzData[factor_vars] <- lapply(alzData[factor_vars], as.factor)

# Step 5: Convert numeric columns to numeric (ensure you're using the correct column names)
num_vars <- c("Age", "EducationLevel", "BMI", "CognitiveTestScore")  # Updated names
alzData[num_vars] <- lapply(alzData[num_vars], as.numeric)

# Step 6: Create squared terms (with the updated names)
alzData$Age2 <- alzData$Age^2
alzData$BMI2 <- alzData$BMI^2
alzData$CognitiveTestScore2 <- alzData$CognitiveTestScore^2
alzData$EducationLevel2 <- alzData$EducationLevel^2

# Step 7: Handle categorical variables and factor conversion for categorical columns
alzData[sapply(alzData, is.character)] <- lapply(alzData[sapply(alzData, is.character)], as.factor)
# Order matters for some categorical variables
alzData$AlcoholConsumption <- factor(alzData$AlcoholConsumption, order = TRUE, levels = c("Never", "Occasionally", "Regularly"))
alzData$SmokingStatus <- factor(alzData$SmokingStatus, order = TRUE, levels = c("Never", "Former", "Current"))
# Ensure that numeric columns are converted to numeric
alzData[sapply(alzData, is.numeric)] <- lapply(alzData[sapply(alzData, is.numeric)], as.numeric)
```

\newpage

## Exploratory Data Analysis

```{r simple plot,echo=FALSE, fig.width=4, ,fig.height=4,fig.show='hold'}
plot(alzData$AlzheimerDiagnosis,main="Postive and negative \n diagnosis proportion",ylab=c("Frequency count of diagnosis"),xlab=c("Diagnosed with Alzheimer's"))
# allele pop frqeuncy vs ours in the study , alz.org, 
minAge<-min(alzData$Age)
maxAge<-max(alzData$Age)
meanAge<-mean(alzData$Age)
countrySummary<-summary(alzData$Country)
Famsummary<-summary(alzData$FamilyHistory)
GenriskSummary<-summary(alzData$GeneticRisk)
totalCount<-nrow(alzData)
```


*Some interesting statistics about our data:*

The total sample size is `r totalCount`

The minimum of age is : `r minAge`

The maximum of age is : `r maxAge`

The mean of age is : `r meanAge`

Countries range in sample size collected: `r countrySummary[1]`,`r countrySummary[2]`

Individuals with reported family history of Alzheimer's responded: `r Famsummary` *(no,yes)*

Individuals tested for the genetic risk factor of the APOE-$\epsilon4$ allele displayed: `r GenriskSummary` *(no,yes)*

\newpage

## Analysis Methods

We will be creating models, comparing similar models AIC criterion and calculating log-odds.

### AIC: 

Akaike Information Criterion (AIC) is a measure used to compare models by balancing goodness of fit and model complexity. It helps identify the best model without overfitting.

How AIC Works:
Formula: $AIC=2k-\ln(L)$

where: k = number of parameters in the model and L = likelihood of the model

In simple terms a lower AIC is better therefore A smaller AIC value suggests a better model.

We can use AIC then to compare models, we fitting multiple models and computing then comparing who has the lowest AIC.

We assume the binary outcome of positive diagnosis should follow a binomial distribution,following this assumption we test a logistic model on all variables.

(4.4) *(4)* and *(5)*

### Logistic Regression model on all variables

```{r Full.Log.Reg, include=FALSE}
# Logistic Regression model on all variables
alz_logit_full <- glm(AlzheimerDiagnosis~ .
                  ,data = alzData
                  ,family = binomial)
sum(vif(alz_logit_full) >= 5) #Check if any vifs are >= 5.
#summary(alz_logit_full)
full.model.aic <- alz_logit_full$aic
# Country, Age, FamilyHistory, and GeneticRisk seem to be the only significant variables.
```

First, we identify what variables may have a significant relationship with a positive (or negative) Alzheimer's diagnosis:
 
Utilizing the general linear model regression from the [@MASS] package, a saturated logistic regression model was created to provide benchmark for evaluating the fit for simpler models. The underlying objective is to find the most parsimonious model and identify the variables that explain an Alzheimer's Diagnosis.
 
As many of the variables were converted to factors as a result of them being categorical, the saturated model was 58 predictor variables along with the intercept term. Since there is such a large amount of predictor variables included, we tested for multi-collinearity to see if any of the variables had a significant relationship to any other.
 
 To detect multi-collinearity between variables, we used Variance Inflation Factors (VIF)[@salmerón2018a] to quantify how much variance each of the regression coefficients is inflated by as a result of the linear regression model. Mathematically, the VIF for the $j$-th predictor variable ($X_j$) is calculated as:
 
 $$
 \text{VIF}_j = \frac{1}{1 - R_j^2}
 $$
 
Here, $R_j^2$ represents the coefficient of determination obtained by regressing $X_j$ on all other predictor variables in the model. A higher VIF indicates greater multicollinearity, suggesting that the variable may be linearly dependent on other variables. A commonly agreed upon VIF that would indicate significant multicollinearity is 10. The result of this testing yielded no VIF value greater than 2 in the saturated model, and as such it is concluded that there exists little to no multicollinearity among all of the predictors in the dataset.
 
 ### Variable Selection
 
To find a more parsimonious model, three stepwise variable selection methods are performed – each using the same function "step" from the [@MASS] package. The stepwise algorithm iterates through the variables of the model passed into the function along with the direction and extracts and compares the Akaike's Information Criterion (AIC) for each variable to determine an optimal model fitted by maximum likelihood. This value is ultimately used to compare each model of the dataset to analyze how well a model fits. Mathematically, the AIC in each step in the function is calculated as:

(4.4) *(4)*
 
 $$
 2n_{par}-2log(\hat{L})
 $$
 
where $n_{par}$ represents the number of parameters in the fitted model. Upon finding an optimal model, that AIC value is recorded in the [@ModelComparison] Table found in the conclusion. For the purposes of finding a parsimonious model, we tested all three directions, namely: "both", "forward", and "backward", as we did not know if they would be identical or not. Results concluded that the three stepwise variable selection directions yield the same model. Because of this, only the "forward" selection model will be referenced going forward, for no other reason than it was the fastest in computation-time.


```{r Step.fwd, echo=FALSE}
#| cache: true
# Intercept model
alz_logit_int <- glm(AlzheimerDiagnosis~1,data=alzData,family="binomial")
 
# Step-wise feature selection ("forward" direction)
step.fwd.model<- step(alz_logit_int, direction = "forward", 
                      scope = list(lower = alz_logit_int, upper = alz_logit_full),
                      trace = F) 
vif(step.fwd.model) # Check if any vifs exceed 5 in forward selection model. None exceed 2
summaryofstep<-summary(step.fwd.model)
```

`r summaryofstep[1]`

Once again, the VIF function is used to ensure that no multicollinearly was introduced with the reduction of predictor variables. There was no significant VIF values.

The glm model as determined by the stepwise forward selection method suggests that variables: "Country", "Age", "FamilyHistory", "GeneticRisk", and "UrbanvsRuralLiving" variables provide the "best" model. The first four listed make some intuitive sense from our understandings of Alzheimer's Disease, however "UrbanvsRuralLiving" is somewhat strange intuitively. In addition, the algorithm that determined this optimal model included this variable even though it showed a borderline significant p-value of 0.0961 which is not less than 0.05 as is the common practice. To further investigate this anomaly, deviance testing was performed to validate if this variable can be removed without significantly affecting the models fit.

*TODO:Introduce deviance testing method. What is it, what does it look like, why I am using it, what are my results.*

Deviance Test: Compare the best feature selected reduced model to one-variable removed models. If removing a variable results in a very large increase in deviance (and a very small p-value), it suggests that the variable plays an important role in the model fit.

Results of deviance tests suggest that removing the Urban vs Rural Living variable from the model would not significantly affect the model fit supported by its p-value being 0.096, which is greater than 0.05.

Conclusions made from test: the best reduced model from feature selection methods could be further simplified without significantly affecting the model.

```{r Step.fwd.deviance.tests, echo=FALSE}
## Fit reduced models (each with one variable removed)

no_country.model <- glm(AlzheimerDiagnosis ~ 
                          Age + FamilyHistory + GeneticRisk + UrbanvsRuralLiving, 
                        data = alzData, family = binomial)

no_age.model <- glm(AlzheimerDiagnosis ~ 
                      Country + FamilyHistory + GeneticRisk + UrbanvsRuralLiving,
                    data = alzData, family = binomial)

no_famHist.model <- glm(AlzheimerDiagnosis ~ 
                          Country + Age + GeneticRisk + UrbanvsRuralLiving, 
                        data = alzData, family = binomial)

no_genRisk.model <- glm(AlzheimerDiagnosis ~ 
                          Country + Age + FamilyHistory + UrbanvsRuralLiving, 
                        data = alzData, family = binomial)

no_urbanvsRural.model <- glm(AlzheimerDiagnosis ~ 
                          Country + Age + FamilyHistory + GeneticRisk, 
                        data = alzData, family = binomial)


## Deviance test: Compare the best feature selected reduced model to one-variable removed models
## If removing a variable results in a very large increase in deviance (and a very small p-value), it suggests that the variable plays an important role in the model fit.
sig.no_country <- pchisq(no_country.model$deviance-step.fwd.model$deviance,
                         df=no_country.model$df.residual-step.fwd.model$df.residual,
                         lower.tail = FALSE)

sig.no_age <- pchisq(no_age.model$deviance-step.fwd.model$deviance,
                     df=no_age.model$df.residual-step.fwd.model$df.residual,
                     lower.tail = FALSE)

sig.no_famHist <- pchisq(no_famHist.model$deviance-step.fwd.model$deviance,
                         df=no_famHist.model$df.residual-step.fwd.model$df.residual,
                         lower.tail = FALSE)

sig.no_genRisk <- pchisq(no_genRisk.model$deviance-step.fwd.model$deviance,
                         df=no_genRisk.model$df.residual-step.fwd.model$df.residual,
                         lower.tail = FALSE)

sig.no_urbanvsRural <- pchisq(no_urbanvsRural.model$deviance-step.fwd.model$deviance,
                         df=no_urbanvsRural.model$df.residual-step.fwd.model$df.residual,
                         lower.tail = FALSE)

# Create a table of results
results <- data.frame(
  `Variable Removed` = c("Country", "Age", "Family History", "Genetic Risk", "Urban vs Rural Living"),
  `Degree Freedom Difference` = c(no_country.model$df.residual-step.fwd.model$df.residual,
                             no_age.model$df.residual-step.fwd.model$df.residual,
                             no_famHist.model$df.residual-step.fwd.model$df.residual,
                             no_genRisk.model$df.residual-step.fwd.model$df.residual,
                             no_urbanvsRural.model$df.residual-step.fwd.model$df.residual),
  `P-value` = c(sig.no_country, sig.no_age, sig.no_famHist, sig.no_genRisk, sig.no_urbanvsRural)
)

#Title = "Significance tests for the reduced model"
kable(results,digits=4)
```

Fitting suggested reduced model and exploring it's results.

```{r Reduced.model.check, echo=FALSE}
## Fit reduced model
alz_logit_reduced <- glm(AlzheimerDiagnosis~ 
                   Country+ Age+ FamilyHistory+ GeneticRisk
                  ,data = alzData
                  ,family = binomial)
#summary(alz_logit_reduced)
alz_logit_red.aic <- alz_logit_reduced$aic
alz_logit_red.aic
plot(alz_logit_reduced, which = 2)
```

The residuals in the QQ plot are heavily skewed to right, indicating a skewed normal or non normal distribution. Instead of general linear models, we introduce general additive models for their smoothing attributes specifically on our age predictor which seeminly follows a distinct step-wise looking function pattern.

Introduce GAM model with splines method. What is it, what does a general GAM method look like (in latex), might also have to give a blurb about splines and how they fit in GAM models.

Fitting a GAM model and putting splines on Age since it's the only variable that is not categorical. Discuss results.

Splines on Age show high significance according to p-value. Gam check plots shows bimodal residuals, which eliminated the skewness from the previous model

```{r GAM.model, eval=FALSE, include=FALSE}
alz_logit_gam<- gam(AlzheimerDiagnosis~ 
                    Country+ s(Age, knots = c(65,65,75,75))+ FamilyHistory+ GeneticRisk
                  ,data = alzData
                  ,family = binomial)
summary(alz_logit_gam)
# Splines on Age show high significance according to p-value.
alz_logit_gam.aic <- alz_logit_gam$aic
alz_logit_gam.aic

gam.check(alz_logit_gam)
#gam check plots shows bimodal, normal residuals, which eliminated the skewness from the previous model
```

*TODO Introduce log odds method; what is it (show a formula in latex), why are we using it, interpret our results. Cite a textbook.*


```{r Log.odds, eval=FALSE, include=FALSE}
#Calculate log-odds ratios for the reduced models to compare the magnitudes across predictors to see which ones have the strongest change in odds
alz_reduced.log_odds_ratio <- exp(coef(alz_logit_reduced))
alz_reduced.log_odds_ratio

alz_gam.log_odds_ratio <- exp(coef(alz_logit_gam))
alz_gam.log_odds_ratio
#Interpret the log odds ratio of the 2 models by comparing the magnitudes. Put into a table for clarity, sorted desc

# Convert to data frames
df_reduced <- data.frame(Predictor = names(alz_reduced.log_odds_ratio), LogOdds_Reduced = alz_reduced.log_odds_ratio)
df_gam <- data.frame(Predictor = names(alz_gam.log_odds_ratio), LogOdds_GAM = alz_gam.log_odds_ratio)

# Merge both data frames by Predictor (outer join to keep all predictors)
log_odds_table <- full_join(df_reduced, df_gam, by = "Predictor")

# Sort by the strongest log-odds in the reduced model
log_odds_table <- log_odds_table %>% arrange(desc(abs(LogOdds_GAM)))

# Display table
kable(log_odds_table, caption = "Comparison of Log-Odds Ratios Between Reduced Models")
```

We can conclude from this comparison of log odds that:...

## Age variable Analysis:

```{r AGE_diagnosis_plot, echo=FALSE,fig.cap="Count of 'Yes' Alzheimer's Diagnosis and Total Sample Size by Age"}
# Count the number of 'Yes' diagnoses and total samples per age
alzData_countage <- alzData %>%
  group_by(Age) %>%
  summarise(
    Yes_Count = sum(AlzheimerDiagnosis == "Yes", na.rm = TRUE),
    Total_Sample_Size = n()
  )
# Combine Yes_Count and Total_Sample_Size for plotting
bar_height <- rbind(alzData_countage$Yes_Count, alzData_countage$Total_Sample_Size)
# Create the bar plot
barplot(bar_height, beside = TRUE, col = c("purple", "red"), 
        names.arg = alzData_countage$Age,
        xlab = "Age", 
        ylab = "Positive diagnosis count and \n total sample size aggregated for ages",
        legend.text = c("Positive Alzheimer's diagnosis", "Total Sample Size"), 
        args.legend = list(title = "Legend",x="bottomright") )

```
We can find this almost step wise trend relating to total diagnosis's counted for each age group (50-65, 65-75,75-100)

```{r step-function, echo=FALSE, fig.show='hold', fig.width=3.5, warning=FALSE, fig.height=3.5}
#Ask john about plots here
table(cut(alzData_countage$Age, breaks = c(min(alzData_countage$Age), 65, 75, max(alzData_countage$Age))))
alzData_countage$Age_Cut <- cut(alzData_countage$Age, breaks = c(min(alzData_countage$Age), 65, 75, max(alzData_countage$Age)))
fit_step <- lm(Yes_Count ~ Age_Cut, data = alzData_countage)
print(coef(summary(fit_step)))
agelims<-range(alzData_countage$Age)
age_grid = seq(from = min(agelims), to = max(agelims))
# Predict the value of the generated ages, returning the standard error using se = TRUE
preds = predict(fit_step, newdata = alzData_countage, se = TRUE)
# Compute error bands (2*SE)
se_bands = cbind("upper" = preds$fit+2*preds$se.fit, 
                 "lower" = preds$fit-2*preds$se.fit)
# Plot
ggplot() +
  geom_point(data = alzData_countage, aes(x = Age, y = Yes_Count)) +
  geom_line(aes(x = age_grid, y = preds$fit), color = "#0000FF") +
  geom_ribbon(aes(x = age_grid, 
                  ymin = se_bands[,"lower"], 
                  ymax = se_bands[,"upper"]), 
              alpha = 0.3) +
  xlim(agelims) +
  labs(title = "Step Function")
hist(residuals(fit_step), main = "Histogram of Residuals", breaks = 20)
qqnorm(residuals(fit_step))
qqline(residuals(fit_step), col = "red")
#plot(fit_step)
```


To create this step function we distinguished groups: 50-65, 65-75,75-100 for age.

## B-Spline Age Model

Using a B-spline model, we can also model the relationship between the count of positive diagnosis and age ranges. From the visual interpretation earlier, we can approximate knots at (65 and 75).

We have a 1st and 3rd degree b-spline on the same knots,

1st degree B-splines are computed by:

$$ B_{i,1}(x)=\frac{(x-t_i)}{t_{i+1}-t_i} 1_{x\exists [t_i,t_{i+1}]} (x) +\frac{(t_{i+2}-x)}{t_{i+2}-t_{i+1}} 1_{x\exists [t_{i+1},t_{i+2}]} (x), i=-1,2,...,k $$ 

(9.5) *(4)*

where as 3rd degree b-splines are calculated recursively using:

$$ B_{i,p}(x)=\frac{(x-t_i)}{t_{i+p}-t_i} B_{i,p-1} (x) +\frac{(t_{i+2}-x)}{t_{i+2}-t_{i+1}} B_{i+1,p-1}  (x), i=-1,2,...,k $$ 

(9.6) *(4)*


```{r Bspline, fig.height=3.5, fig.width=3.5,fig.show='hold', echo=FALSE}
# spline model, deg 1 and 3
bsplineProj<-lm(alzData_countage$Yes_Count~bs(alzData_countage$Age,knots=c( 65,65,75,75),degree=1),data=alzData_countage)
bsplineProjDeg3<-lm(alzData_countage$Yes_Count~bs(alzData_countage$Age,knots=c( 65,65,75,75),degree=3),data=alzData_countage)
plot(alzData_countage$Yes_Count~alzData_countage$Age, data = alzData_countage,pch=16,main="bspline with knots 65 and 75",ylab="Positive diagnosis",xlab="Age")
pred <- predict(bsplineProj, newdata = data.frame(alzData_countage$Age), interval = "prediction")
y_fit <- pred[, "fit"]
upper_pred <- pred[, "upr"]
lower_pred <- pred[, "lwr"]
with(data.frame(alzData_countage$Age, y_fit), lines(alzData_countage$Age, y_fit, col=4, lwd=2))
with(data.frame(alzData_countage$Age, upper_pred), lines(alzData_countage$Age, upper_pred, col="red", lty=2))
with(data.frame(alzData_countage$Age, lower_pred), lines(alzData_countage$Age, lower_pred, col="red", lty=2))
plot(alzData_countage$Yes_Count~alzData_countage$Age, data = alzData_countage,pch=16,main="bspline with knots 65 and 75, deg 3",ylab="Positive diagnosis",xlab="Age")
pred1 <- predict(bsplineProjDeg3, newdata = data.frame(alzData_countage$Age), interval = "prediction")
y_fit1 <- pred1[, "fit"]
upper_pred1 <- pred1[, "upr"]
lower_pred1 <- pred1[, "lwr"]
with(data.frame(alzData_countage$Age, y_fit), lines(alzData_countage$Age, y_fit1, col=4, lwd=2))
with(data.frame(alzData_countage$Age, upper_pred1), lines(alzData_countage$Age, upper_pred1, col="red", lty=2))
with(data.frame(alzData_countage$Age, lower_pred1), lines(alzData_countage$Age, lower_pred1, col="red", lty=2))
#plot(bsplineProjDeg3)
```

The splines model of third degree seems to capture the relationship between age and number of diagnosis well, so it may be better to imagine age as a category in response to diagnosis count.


Since our min and max of age in the data is `r minAge`, `r maxAge`

```{r ANOVA, echo=FALSE}
alzData <- alzData %>%
  mutate(AgeCategory = case_when(
    Age >= 50 & Age <= 65 ~ "50-65",
    Age > 65 & Age <= 75 ~ "65-75",
    Age > 75 & Age <= 100 ~ "75-100",
    TRUE ~ "Other"
  ))
# Calculate diagnosis rate (proportion of Yes diagnoses)
alzData <- alzData %>%
  mutate(diagnosis_rate = ifelse(AlzheimerDiagnosis == "Yes", 1, 0))
# Perform ANOVA for diagnosis rate by Country, Gender, FamilyHistory, and AgeCategory
anova_result <- aov(diagnosis_rate ~ Country + Gender + FamilyHistory + AgeCategory, data = alzData)
# Summarize the ANOVA result
print(summary(anova_result))

```

```{r Plot_country_prop, echo=FALSE}
# Calculate the average diagnosis rate for each combination of Country and Age Category
alzData_avg <- alzData %>%
  group_by(Country, AgeCategory) %>%
  summarise(diagnosis_rate = mean(AlzheimerDiagnosis == "Yes", na.rm = TRUE)) %>%
  ungroup()  # Ensure that we don't retain grouping in the data
# Now, create the interaction plot
ggplot(alzData_avg, aes(x = AgeCategory, y = diagnosis_rate, color = Country, group = Country)) +
  geom_point(size = 3) +  # Add points for each diagnosis rate
  geom_line() +  # Add a line to show the trend within each country
  labs(title = "Diagnosis Rate by Age Category for all Countries",
       x = "Age Category", 
       y = "Average Diagnosis Rate") +
  theme_minimal() +  # Clean plot theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```


Now to simplify we display only the top and bottom averages overall.

```{r All Countries plot, echo=FALSE}
# Step 1: Calculate the diagnosis rate for each combination of Country and Age Category
alzData_avg <- alzData %>%
  group_by(Country, AgeCategory) %>%
  summarise(diagnosis_rate = mean(AlzheimerDiagnosis == "Yes", na.rm = TRUE)) %>%
  ungroup()

# Step 2: Identify the top and bottom countries based on the average diagnosis rate
country_avg_rates <- alzData_avg %>%
  group_by(Country) %>%
  summarise(avg_diagnosis_rate = mean(diagnosis_rate, na.rm = TRUE)) %>%
  arrange(desc(avg_diagnosis_rate))

# Get top and bottom countries
top_country <- country_avg_rates$Country[1]
bottom_country <- country_avg_rates$Country[nrow(country_avg_rates)]

# Step 3: Filter the data for only the top and bottom countries
alzData_top_bottom <- alzData_avg %>%
  filter(Country %in% c(top_country, bottom_country))

# Step 4: Create the plot for the top and bottom countries
ggplot(alzData_top_bottom, aes(x = AgeCategory, y = diagnosis_rate, color = Country, group = Country)) +
  geom_point(size = 3) +  # Add points for each diagnosis rate
  geom_line() +  # Add a line to show the trend within each country
  labs(title = "Diagnosis Rate by Age Category for Top and Bottom Country",
       x = "Age Category", y = "Average Diagnosis Rate") +
  theme_minimal() +  # Clean plot theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```


```{r Plotofimportant,echo=FALSE}
# Create a contingency table
alz_table <- table(alzData$FamilyHistory, alzData$AlzheimerDiagnosis)
# Perform Chi-Square Test
chi_test <- chisq.test(alz_table)
# Print results
print(chi_test)
```

Motivation for quad

## Quadratic and log models:

```{r ChangeVarEmily, echo=FALSE}
data<-alzData
colnames(data)[colnames(data) == "Family.History.of.Alzheimer.s"] <- "FamilyHistory"
colnames(data)[colnames(data) == "Genetic.Risk.Factor..APOE.ε4.allele."] <- "GeneticRisk"
colnames(data)[colnames(data) == "Alzheimer.s.Diagnosis"] <- "AlzheimerDiagnosis"
colnames(data)[colnames(data) == "Physical.Activity.Level"] <- "PhysicalActivityLevel"
colnames(data)[colnames(data) == "Smoking.Status"] <- "SmokingStatus"
colnames(data)[colnames(data) == "Cholesterol.Level"] <- "CholesterolLevel"
colnames(data)[colnames(data) == "Cognitive.Test.Score"] <- "CognitiveTestScore"
colnames(data)[colnames(data) == "Sleep.Quality"] <- "SleepQuality"
colnames(data)[colnames(data) == "Air.Pollution.Exposure"] <- "AirPollutionExposure"
colnames(data)[colnames(data) == "Marital.Status"] <- "MaritalStatus"
colnames(data)[colnames(data) == "Social.Engagement.Level"] <- "SocialEngagementLevel"
colnames(data)[colnames(data) == "Stress.Levels"] <- "StressLevels"
colnames(data)[colnames(data) == "Education.Level"] <- "EducationLevel"
colnames(data)[colnames(data) == "Alcohol.Consumption"] <- "AlcoholConsumption"
colnames(data)[colnames(data) == "Depression.Level"] <- "DepressionLevel"
colnames(data)[colnames(data) == "Dietary.Habits"] <- "DietaryHabits"
colnames(data)[colnames(data) == "Employment.Status"] <- "EmploymentStatus"
colnames(data)[colnames(data) == "Income.Level"] <- "IncomeLevel"
colnames(data)[colnames(data) == "Urban.vs.Rural.Living"] <- "UrbanvsRuralLiving"

trans_variables <- c("Country", "Gender", "PhysicalActivityLevel", "SmokingStatus", 
                 "AlcoholConsumption", "Diabetes", "Hypertension", "CholesterolLevel", 
                 "FamilyHistory", "DepressionLevel", "SleepQuality", "DietaryHabits", 
                 "AirPollutionExposure", "EmploymentStatus", "MaritalStatus", 
                 "GeneticRisk", "SocialEngagementLevel", "IncomeLevel", "StressLevels", 
                 "UrbanvsRuralLiving", "AlzheimerDiagnosis")
data[trans_variables] <- lapply(data[trans_variables], as.factor)
trans_num_vars <- c("Age", "EducationLevel", "BMI", "CognitiveTestScore")
data[trans_num_vars] <- lapply(data[trans_num_vars], as.numeric)
data$Age2 <- data$Age^2
data$BMI2 <- data$BMI^2
data$CognitiveTestScore2 <- data$CognitiveTestScore^2
data$EducationLevel2 <- data$EducationLevel^2
```

quadratic

```{r Quad model, echo=FALSE}
quadratic_model <- glm(AlzheimerDiagnosis ~ Age + Age2 + BMI + BMI2 +CognitiveTestScore + CognitiveTestScore2 + EducationLevel + EducationLevel2 +PhysicalActivityLevel + SmokingStatus + AlcoholConsumption + Hypertension + Diabetes + CholesterolLevel + FamilyHistory +DepressionLevel + SleepQuality + DietaryHabits + AirPollutionExposure + EmploymentStatus + MaritalStatus +GeneticRisk + SocialEngagementLevel + IncomeLevel + StressLevels +UrbanvsRuralLiving + Gender, family = binomial, data = data)
#AIC(quadratic_model)
#BIC(quadratic_model)
```

logistic

```{r logistic Model, echo=FALSE}
logistic_model <- glm(AlzheimerDiagnosis ~ Age + BMI + CognitiveTestScore + EducationLevel +PhysicalActivityLevel + SmokingStatus + AlcoholConsumption + Hypertension + Diabetes + CholesterolLevel + FamilyHistory +DepressionLevel + SleepQuality + DietaryHabits+ AirPollutionExposure + EmploymentStatus + MaritalStatus +GeneticRisk + SocialEngagementLevel + IncomeLevel + StressLevels + UrbanvsRuralLiving + Gender,family = binomial, data = data)
#AIC(logistic_model)
#BIC(logistic_model)
```

Based on AIC and BIC, the quadratic model's are lower, so its performance better than logistic

```{r FamilyHist, echo=FALSE}
data %>%
  group_by(FamilyHistory) %>%
  summarise(rate = mean(AlzheimerDiagnosis == "Yes", na.rm = TRUE))

ggplot(data, aes(x = FamilyHistory, fill = AlzheimerDiagnosis)) +
  geom_bar(position = "fill") +
  ylab("Proportion of Diagnosed") +
  labs(title = "Alzheimer's Diagnosis by Family History")
```

As can be seen from the chart, among those without a Family History, 36.8% were diagnosed with Alzheimer, while among those with a Family History, 51.9% were diagnosed with Alzheimer.Therefore, Family History is an important indicator for judging Alzheimer, and people with a Family History are more likely to suffer from Alzheimer.

```{r GeneticRisk, echo=FALSE}
data %>%
  group_by(GeneticRisk) %>%
  summarise(rate = mean(AlzheimerDiagnosis == "Yes", na.rm = TRUE))
ggplot(data, aes(x =GeneticRisk, fill = AlzheimerDiagnosis)) +
  geom_bar(position = "fill") +
  ylab("Proportion of Diagnosed") +
  labs(title = "Alzheimer's Diagnosis by Genetic Risk")
```

Among people without genetic risk, the proportion of people diagnosed with Alzheimer's is 36.6%. In some people, this probability is 60.6%. As above, this is also a very important indicator, and people with genetic risk have a higher probability of developing Alzheimer's.

```{r GLM_int, echo=FALSE}
model_af <- glm(AlzheimerDiagnosis ~ Age * FamilyHistory,data = data, family = binomial)
#summary(model_af)
#AIC(model_af)
cv_model_af <- cv.glm(data, model_af, K = 10)
cv_model_af$delta
```
The Interaction between Age and Family is significant because p-value is 0.012 which is smaller than 0.05

## Gam with CV

```{r GAM, echo=FALSE}
gam_af <- gam(AlzheimerDiagnosis ~ s(Age, by = FamilyHistory, k = 5) + FamilyHistory,data = data, family = binomial)
plot(gam_af, pages = 1, se = TRUE)
AIC(gam_af)
data$AlzheimerDiagnosis <- ifelse(data$AlzheimerDiagnosis == "Yes", 1, 0)
cv_gam_af <- cv.glm(data, gam_af, K = 10)
cv_gam_af$delta
```

No FH:50-60 stable, 60-80 increase sharply; 80+ stable Yes FH:50-60 higher than NO FH, 60-80 same as NO FH

```{r ModelCompare, echo=FALSE}
model_comparison <- data.frame(
  Model = c("GLM_af", "GAM_af"),
  AIC = c(AIC(model_af), AIC(gam_af)),
  CV = c(cv_model_af$delta[1], cv_gam_af$delta[1])
)
model_comparison

```

Alzheimer's risk increases with the age. And risk becomes higher with family history. There is evidence of an interaction between age and family history on Alzheimer’s diagnosis

```{r AIC compare, echo=FALSE}
# Create an empty data frame to store AIC values
aic_comparison <- data.frame(Model = character(), AIC = numeric(), stringsAsFactors = FALSE)
# Add AIC values for each model
aic_comparison <- rbind(aic_comparison, data.frame(Model = "GAM Interaction", AIC = AIC(gam_af)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "GLM Model", AIC = AIC(model_af)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "logistic model", AIC = AIC(logistic_model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "log reduced", AIC = AIC(alz_logit_reduced)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "quadratic model", AIC = AIC(quadratic_model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "Step forward model", AIC = AIC(step.fwd.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "No Age Log", AIC = AIC(no_age.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "No family hist log", AIC = AIC(no_famHist.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "No genRisk log", AIC = AIC(no_genRisk.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "No urban/rural log", AIC = AIC(no_urbanvsRural.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "No country log", AIC = AIC(no_country.model)))
aic_comparison <- rbind(aic_comparison, data.frame(Model = "full log", AIC = AIC(alz_logit_full)))
#aic_comparison <- rbind(aic_comparison, data.frame(Model = "Gam log", AIC = AIC(alz_logit_gam)))

# View the AIC comparison table
print(aic_comparison)
```

```{r}
#TODO area
# Create age categories
#TODO add t-tests for genetic risk and family history
#TODO look into rate
#significance test on japan vs india
#significance on age groups
#predict based on model
#log odds
#TODO caching chunks, formatting, remove MLR
#Get ride /format summaries
##For final report, for each method: include what a method is and if not mentioned in class - a citation from some textbook that the method is found, how it is done (latex of formula or process), reasoning for the method, and discuss the results.
```

Bibliography:

#### <https://www.nia.nih.gov/health/alzheimers-causes-and-risk-factors/alzheimers-disease-genetics-fact-sheet> 
*(1)*

#### <https://www.science.smith.edu/~jcrouser/SDS293/labs/lab12-r.html> 
*(2)*

#### <https://www.nia.nih.gov/health/alzheimers-and-dementia/understanding-different-types-dementia> 
*(3)*

#### <Linear Regression and Generalized Linear Models W. John Braun University of British Columbia> 
*(4)*

#### <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/AIC>

*(5)*

 <!--# VIF Citation -->
 
 [@salmerón2018a]
 
 [@MASS]