---
title: "Investigating Factors Associated with Alzheimer's Diagnosis"
subtitle: 
author: "Ben daSilva Curiel 35207323 , Dylan Forsythe 47473699 , Yiran Wang 55134043"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{tcolorbox}
output: pdf_document
bibliography: references.bib
---

## Introduction

Alzheimer's disease is a progressive brain disease and the most common cause and type of dementia . Dementia causes a decline in memory and cognitive abilities, and in severe cases can even affect basic functions of daily life. The disease also shortens the lifespan of patients and places long-term physical and mental burdens on patients and their families. More than half of all dementia patients in the world suffer from Alzheimer's disease, and the disease is currently incurable, the interventions only slowing its progression [@2024alz2024].

As the population ages, the incidence of Alzheimer's disease continues to rise. It is estimated that by 2030, there will be 78 million people suffering from dementia worldwide, and by 2050, there will be 139 million patients. This will put huge pressure on the medical system and social resources [@2024alz2024]. Therefore, early identification and judgment of disease risk is not only crucial for patients and their families to better deal with the disease, but also contributes to a more rational allocation of medical resources in the society. Given its widespread impact, prevention and early identification of the disease are particularly important.

This article selects a data set on Alzheimer's Diagnosis' [@ankit2025] on Kaggle.com on Feb. 6, 2025 to analyze. This data set was created by Ankush Panday. This data was released under the MIT license. According to the Kaggle webpage, Panday utilized web scraping techniques to collect the data from two health reports in 2023 and will be updated on an annual basis. This means that the data may come from public health databases, research papers, or online resources of government or medical institutions, rather than from surveys or experiments. There is no missing data or duplicate values found.

## Motivation

The motivation for this research is preventative, to help identify possible lifestyle and genetic indicators of a positive Alzheimer's diagnosis and understand the despairing reality of the disease. By researching age-related brain changes, researchers can better distinguish between normal aging and pathological changes, which is essential for accurate diagnosis and understanding disease mechanisms. Additionally, age influences how individuals respond to medications and treatments, highlighting the importance of studying age-related differences in treatment responses to develop more personalized therapeutic approaches and optimize care for elderly patients. This research is particularly crucial given the shifting global population structure, where many countries are experiencing an aging population with more elderly individuals than younger ones. As life expectancy increases and birth rates decline, the prevalence of age-related diseases like Alzheimer's is expected to rise, placing a greater burden on healthcare systems and society. Furthermore, studying the impact of age on Alzheimer's could help identify early bio-markers and warning signs, paving the way for better diagnostic tools and enabling earlier interventions, when treatments may be more effective in slowing or preventing disease progression. Addressing the challenges posed by an aging population requires a deeper understanding of how age affects Alzheimer's risk, progression, and treatment outcomes.

```{r Librarysetup, include=FALSE}
library(mgcv)
library(dplyr)
library(ggplot2)
library(splines)
library(car)
library(MASS)
library(parallel)
library(tidyr)
library(boot)
library(DAAG)
library(caret)
library(knitr)
library(kableExtra)
library(broom)
library(np)
library(epitools)
```

```{r DataImport,include=FALSE}
setwd("/Users/Bendasilva/Desktop/410_Project")

# NOTE TO BEN: DONT CHANGE THIS PART ON YOURS. THIS IS JUST SO I CAN KNIT
#alzData<-read.csv("alzheimers_prediction_Dataset.csv",header = TRUE,sep=",")
alzData<-read.csv("alzheimers_prediction_Dataset.csv",header = TRUE,sep=",")
```

```{r ColNameChange,include=FALSE}
## The conversion
# Step 1: Rename columns to match the simplified names
colnames(alzData)[colnames(alzData) == "Family.History.of.Alzheimer.s"] <- "FamilyHistory"
colnames(alzData)[colnames(alzData) == "Genetic.Risk.Factor..APOE.ε4.allele."] <- "GeneticRisk"
colnames(alzData)[colnames(alzData) == "Alzheimer.s.Diagnosis"] <- "AlzheimerDiagnosis"
colnames(alzData)[colnames(alzData) == "Physical.Activity.Level"] <- "PhysicalActivityLevel"
colnames(alzData)[colnames(alzData) == "Smoking.Status"] <- "SmokingStatus"
colnames(alzData)[colnames(alzData) == "Cholesterol.Level"] <- "CholesterolLevel"
colnames(alzData)[colnames(alzData) == "Cognitive.Test.Score"] <- "CognitiveTestScore"
colnames(alzData)[colnames(alzData) == "Sleep.Quality"] <- "SleepQuality"
colnames(alzData)[colnames(alzData) == "Air.Pollution.Exposure"] <- "AirPollutionExposure"
colnames(alzData)[colnames(alzData) == "Marital.Status"] <- "MaritalStatus"
colnames(alzData)[colnames(alzData) == "Social.Engagement.Level"] <- "SocialEngagementLevel"
colnames(alzData)[colnames(alzData) == "Stress.Levels"] <- "StressLevels"
colnames(alzData)[colnames(alzData) == "Education.Level"] <- "EducationLevel"
colnames(alzData)[colnames(alzData) == "Alcohol.Consumption"] <- "AlcoholConsumption"
colnames(alzData)[colnames(alzData) == "Depression.Level"] <- "DepressionLevel"
colnames(alzData)[colnames(alzData) == "Dietary.Habits"] <- "DietaryHabits"
colnames(alzData)[colnames(alzData) == "Employment.Status"] <- "EmploymentStatus"
colnames(alzData)[colnames(alzData) == "Income.Level"] <- "IncomeLevel"
colnames(alzData)[colnames(alzData) == "Urban.vs.Rural.Living"] <- "UrbanvsRuralLiving"

# Step 2: Update factor_vars list to use the new column names
factor_vars <- c("Country", "Gender", "PhysicalActivityLevel", "SmokingStatus",
                 "AlcoholConsumption", "Diabetes", "Hypertension", "CholesterolLevel", 
                 "FamilyHistory", "DepressionLevel", "SleepQuality", "DietaryHabits", 
                 "AirPollutionExposure", "EmploymentStatus", "MaritalStatus", 
                 "GeneticRisk", "SocialEngagementLevel", "IncomeLevel", "StressLevels", 
                 "UrbanvsRuralLiving", "AlzheimerDiagnosis")

# Step 3: Check if all factor variables exist in the data
missing_columns <- setdiff(factor_vars, colnames(alzData))
if(length(missing_columns) > 0) {
  stop(paste("Missing columns:", paste(missing_columns, collapse = ", ")))
}

# Step 4: Convert the specified columns to factors (after renaming)
alzData[factor_vars] <- lapply(alzData[factor_vars], as.factor)

# Step 5: Convert numeric columns to numeric (ensure you're using the correct column names)
num_vars <- c("Age", "EducationLevel", "BMI", "CognitiveTestScore")  # Updated names
alzData[num_vars] <- lapply(alzData[num_vars], as.numeric)

# Step 6: Handle categorical variables and factor conversion for categorical columns
alzData[sapply(alzData, is.character)] <- lapply(alzData[sapply(alzData, is.character)], as.factor)
# Order matters for some categorical variables
alzData$AlcoholConsumption <- factor(alzData$AlcoholConsumption, order = TRUE, levels = c("Never", "Occasionally", "Regularly"))
alzData$SmokingStatus <- factor(alzData$SmokingStatus, order = TRUE, levels = c("Never", "Former", "Current"))
# Ensure that numeric columns are converted to numeric
alzData[sapply(alzData, is.numeric)] <- lapply(alzData[sapply(alzData, is.numeric)], as.numeric)

# I had to add this at the beginning because my analysis is based on the original data, without adding columns like AgeCategory, and the quadratic columns
alzDataDylan <- alzData
alzDataEmily <- alzData
```

## Exploratory Data Analysis

```{r simple plot,echo=FALSE, fig.width=4, ,fig.height=4,fig.show='hold', cache=TRUE}
# allele pop frqeuncy vs ours in the study , alz.org, 
minAge<-min(alzData$Age)
maxAge<-max(alzData$Age)
meanAge<-mean(alzData$Age)
countrySummary<-summary(alzData$Country)
Famsummary<-summary(alzData$FamilyHistory)
GenriskSummary<-summary(alzData$GeneticRisk)
totalCount<-nrow(alzData)
```

The data set contains 74,283 records and 25 variables, and the intended purpose is to study the factors associated with an Alzheimer's diagnosis. It includes patients from 20 distinct countries and includes demographic information, health status, lifestyle, socioeconomic background, and genetic factors. Variable types include numerical variables (such as age, BMI, cognitive test scores, education level) and categorical variables (such as gender, smoking status, diabetes, hyper-tension, etc.). Some variables have multi-level classification, for example, 'DepressionLevel' is divided into "Low", "Medium", "High", and could be ordered, and some factors are of a binary classification (such as "Yes" or "No" that an individual has diabetes). The distribution of the "Alzheimer's Diagnosis" column is assumed to be of a binomial distribution because of the binary responses. The following table provides a brief summary of the numerical variables in the data set:

```{=tex}
\begin{tcolorbox}[colframe=black, colback=gray!10, coltitle=black, title=Interesting statistics:]
The total sample size is `r totalCount` individuals, the minimum age of these individuals is: `r minAge` while the maximum age is: `r maxAge`. The mean age is: `r round(meanAge, 3)`

Individuals with reported family history of Alzheimer's responded:
(`r Famsummary`)[(No, Yes) respectively]

Individuals tested for the genetic risk factor of the APOE-$\epsilon4$ allele displayed: 
(`r GenriskSummary`) [(No, Yes) respectively]
\end{tcolorbox}
```
The genetic risk factor mentioned above represents the gene known as apolipoprotein E (APOE) and is known within the research field of dementia to be a leading identification factor for Alzheimer's. This gene is part of a process in making proteins that helps transport cholesterol and other types of fats within the bloodstream. Problems in the gene's expression are well documented in the contributions of development of Alzheimer's. [@benkhedher2021]. The values for the "Genetic Risk" column are binary and represent the presence of this gene in an individual.

```{r FamilyHist,fig.cap="Alzheimer's Diagnosis by Family History\n and Alzheimer's Diagnosis by Genetic Risk",echo=FALSE, fig.show='hold',fig.width=3.5,fig.height=3}

par(mfrow = c(2,2))
# Compute proportions for Family History
family_history_summary <- alzDataEmily %>%
  group_by(FamilyHistory, AlzheimerDiagnosis) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(FamilyHistory) %>%
  mutate(proportion = count / sum(count),
         label = paste0(round(proportion * 100, 1), "%"))  # Format percentage manually

# Plot for Family History with percentages inside bars
ggplot(family_history_summary, aes(x = FamilyHistory, y = proportion, fill = AlzheimerDiagnosis)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = label), position = position_fill(vjust = 0.5), size = 3) +
  ylab("Proportion of Diagnosed") +
  ggtitle("Alzheimer's Diagnosis by \n Family History") +
  theme_minimal()

# Compute proportions for Genetic Risk
genetic_risk_summary <- alzDataEmily %>%
  group_by(GeneticRisk, AlzheimerDiagnosis) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(GeneticRisk) %>%
  mutate(proportion = count / sum(count),
         label = paste0(round(proportion * 100, 1), "%"))  # Format percentage manually

# Plot for Genetic Risk with percentages inside bars
ggplot(genetic_risk_summary, aes(x = GeneticRisk, y = proportion, fill = AlzheimerDiagnosis)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = label), position = position_fill(vjust = 0.5), size = 3) +
  ylab("Proportion of Diagnosed") +
  ggtitle("Alzheimer's Diagnosis by \n Genetic Risk") +
  theme_minimal()
par(mfrow = c(1,1))
```

Among all the variables, the regression coefficients of Family History and Genetic Risk were significant (p values were all less than 0.05), indicating that they had strong predictive power in the model, and were therefore further analyzed separately.

Specifically, the diagnosis rate of Alzheimer's was 51.9% in people with a family history, significantly higher than 36.8% in people without a family history. And in people with genetic risk, the diagnosis rate was 60.6% which was also significantly higher than that of people without genetic risk (36.6%). This result was reflected by the grouped bar chart, showing that the diagnosis rate was higher in people with a family history and genetic risk.

These results show that Family History and Genetic Risk are important influencing factors of Alzheimer's risk, and can be used as potential risk indicators within models to provide support for early screening and risk intervention.

Overall, this data set contains a wide variety of clinical and demographic features that are related to Alzheimer's diagnosis worldwide, making it suitable for exploring potential risk factors for Alzheimer's disease and supports data analysis and modeling simulation.

### Analysis of Suspected Variables

Based on the exploration of the data set and the research from the previously mentioned journals and articles, it is suspected that the possible significant variables associated with Alzheimer's are: Age, Family History, and Genetic Risk. Since the data also includes the Country of origin of each individual, an investigation as to how these variables may differ in each Country was conducted. Age group analysis of Alzheimer's Diagnosis internationally was summarized in the following graphs:

```{r Plot_country_prop, echo=FALSE,fig.show='hold', fig.width=3.5, fig.height=4.5, warning=FALSE}
# Calculate the average diagnosis rate for each combination of Country and Age Category
alzData <- alzData %>%
  mutate(AgeCategory = case_when(
    Age >= 50 & Age <= 65 ~ "50-65",
    Age > 65 & Age <= 75 ~ "65-75",
    Age > 75 & Age <= 100 ~ "75-100",
    TRUE ~ "Other"))
# Calculate diagnosis rate (proportion of Yes diagnoses)
alzData <- alzData %>%
  mutate(diagnosis_rate = ifelse(AlzheimerDiagnosis == "Yes", 1, 0))
alzData_avg <- alzData %>%
  group_by(Country, AgeCategory) %>%
  summarise(diagnosis_rate = mean(AlzheimerDiagnosis == "Yes", na.rm = TRUE), .groups = "drop")

# Now, create the interaction plot
ggplot(alzData_avg, aes(x = AgeCategory, y = diagnosis_rate, color = Country, group = Country)) +
  geom_point(size = 1.5) +  # Add points for each diagnosis rate
  geom_line() +  # Add a line to show the trend within each country
  labs(title = "Diagnosis Rate by Age \nCategory for all Countries",
       x = "Age Category", 
       y = "Average Diagnosis Rate") +
  theme_minimal() +  # Clean plot theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.text = element_text(size = 6),  
    legend.key.size = unit(0.5, "cm"),  
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),  # Add black border around plot
  )
# Step 2: Identify the top and bottom countries based on the average diagnosis rate
country_avg_rates <- alzData_avg %>%
  group_by(Country) %>%
  summarise(avg_diagnosis_rate = mean(diagnosis_rate, na.rm = TRUE)) %>%
  arrange(desc(avg_diagnosis_rate))

# Get top and bottom countries
top_country <- country_avg_rates$Country[1]
bottom_country <- country_avg_rates$Country[nrow(country_avg_rates)]

# Step 3: Filter the data for only the top and bottom countries
alzData_top_bottom <- alzData_avg %>%
  filter(Country %in% c(top_country, bottom_country))

# Step 4: Create the plot for the top and bottom countries
ggplot(alzData_top_bottom, aes(x = AgeCategory, y = diagnosis_rate, color = Country, group = Country)) +
  geom_point(size = 3) +  # Add points for each diagnosis rate
  geom_line() +  # Add a line to show the trend within each country
  labs(title = "Diagnosis Rate by Age Category \nfor Top and Bottom Country",
       x = "Age Category", y = "Average Diagnosis Rate") +
  theme_minimal() +  # Clean plot theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    legend.text = element_text(size = 8),  
    legend.key.size = unit(1, "cm"),  
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),  # Add black border around plot
  )

```

Diagnosis rate does vary country to country per age group with an evidently visual difference between the highest average diagnosed and lowest average diagnosed country (India and Japan, respectively).

Clearly, the rate of Alzheimer's diagnoses increases with age, displaying age as a primary risk factor for the disease. As individuals age, the brain undergoes natural changes, such as shrinkage and the accumulation of plaques and tangles, which can contribute to Alzheimer's [@2024alz2024].

> *An estimated 6.9 million Americans age 65 and older are living with Alzheimer's dementia today. This number could grow to 13.8 million by 2060, barring the development of medical breakthroughs to prevent or cure AD* [@2024alz2024].

\newpage

Aggregating the number of positive diagnosis on each age displays the following information:

```{r AGE_diagnosis_plot, echo=FALSE, warning=FALSE, fig.width=8, fig.height=4}
# Count the number of 'Yes' diagnoses and total samples per age
alzData_countage <- alzData %>%
  group_by(Age) %>%
  summarise(
    Yes_Count = sum(AlzheimerDiagnosis == "Yes", na.rm = TRUE),
    Total_Sample_Size = n(),
        .groups = "drop"  # This removes the grouping after summarising
  )

# Combine Yes_Count and Total_Sample_Size for plotting
bar_height <- rbind(alzData_countage$Yes_Count, alzData_countage$Total_Sample_Size)

par(mar = c(5, 6, 4, 8))  # Increase left margin and reduce right margin
barplot(bar_height, beside = TRUE,
        col = c("skyblue", "pink"),  # Red for positive diagnosis and green for total sample size
        names.arg = alzData_countage$Age, 
        xlab = "Age", 
        main = "Positive Diagnosis Counts \n for each Age",
        #ylab = "Number of Positive Diagnosis", 
        legend.text = c("Positive Alzheimer's Diagnosis", "Total Sample Size \n for age"),
        args.legend = list(title="Legend", 
                          x = "right",
                          inset=c(-0.3,0),
                          cex = 0.6,  # Smaller text size for the legend
                          bty = "y"),  # Remove border around legend
        ylim = c(0, max(bar_height) * 1.2),  
        cex.names = 0.75,  
        cex.axis = 0.9,  
        border = "white",
        width = 0.8)  # Reduce bar width to squeeze the plot
# Reset margins after plotting
par(mar = c(5, 4, 4, 2))
```

An interesting step-function-like relationship was discovered between age and the count of positive diagnoses in the data. This step wise pattern aligns with existing literature, which indicates that the risk of Alzheimer's disease increases significantly after the age of 65. Given this observed trend, B-splines and kernel regression was chosen to approximate the potential non-linear functions, allowing for a more accurate representation of the relationship between age and diagnosis rates.

## B-Spline Age Model

Using a B-spline model, the relationship between the count of positive diagnoses and age ranges can also be modeled. Based on the previous visual interpretation, knots can be approximated at the ages of 66 and 76.

A 1st and 3rd degree B-spline are applied to the same knots.

```{r Bspline,cache=TRUE,echo=FALSE, fig.width=3.5, fig.height=4}
#try non-smooth package from john
bsplineProj<-lm(alzData_countage$Yes_Count~bs(alzData_countage$Age,knots=c( 66,66,76,76),degree=1),data=alzData_countage)
bsplineProjDeg3<-lm(alzData_countage$Yes_Count~bs(alzData_countage$Age,knots=c( 66,66,66,66,76,76,76,76),degree=3),data=alzData_countage)
#summary(bsplineProj)
#summary(bsplineProjDeg3)
```

```{r BsplinePlot, fig.height=3, fig.width=3.25,fig.show='hold', echo=FALSE,cache=TRUE}
plot(alzData_countage$Yes_Count~alzData_countage$Age, data = alzData_countage,pch=16,cex=0.6,main="B-Spline with Knots 65 & 75 \n (Degree 1)",ylab="Positive Diagnosis Count",xlab="Age")
pred <- predict(bsplineProj, newdata = data.frame(alzData_countage$Age), interval = "prediction")
y_fit <- pred[, "fit"]
upper_pred <- pred[, "upr"]
lower_pred <- pred[, "lwr"]
with(data.frame(alzData_countage$Age, y_fit), lines(alzData_countage$Age, y_fit, col=4, lwd=2))
with(data.frame(alzData_countage$Age, upper_pred), lines(alzData_countage$Age, upper_pred, col="red", lty=2))
with(data.frame(alzData_countage$Age, lower_pred), lines(alzData_countage$Age, lower_pred, col="red", lty=2))
plot(alzData_countage$Yes_Count~alzData_countage$Age, data = alzData_countage,pch=16,cex=0.6, main="B-Spline with Knots 65 & 75 \n (Degree 3)",ylab="Positive Diagnosis Count",xlab="Age")
pred1 <- predict(bsplineProjDeg3, newdata = data.frame(alzData_countage$Age), interval = "prediction")
y_fit1 <- pred1[, "fit"]
upper_pred1 <- pred1[, "upr"]
lower_pred1 <- pred1[, "lwr"]
with(data.frame(alzData_countage$Age, y_fit), lines(alzData_countage$Age, y_fit1, col=4, lwd=2))
with(data.frame(alzData_countage$Age, upper_pred1), lines(alzData_countage$Age, upper_pred1, col="red", lty=2))
with(data.frame(alzData_countage$Age, lower_pred1), lines(alzData_countage$Age, lower_pred1, col="red", lty=2))
```

The splines model of third degree seems to capture the relationship between age and number of diagnosis well, so it may be better to imagine age as a category in response to diagnosis count.

## Univariate Kernel Regression Model

Trying a non-parametric model, kernel regression is used to regress the diagnosis count with age. Kernel regression is very flexible, allowing us to model the non-linear jumps in diagnosis per age well. Optimal bandwidth of the kernel was chosen using least-squares cross validation from [@np]. Specifically using the function 'npregbw' to find a optimal bandwidth value of least-squares cross-validation.

```{r Kernel, cache=TRUE,include=FALSE}
# Ensure alzData_countage is loaded in a previous chunk

# Step 1: Select optimal bandwidth using cross-validation
alzDataAge <- alzData_countage
bw <- npregbw(formula = Yes_Count ~ Age, data = alzDataAge)
# Step 2: Perform Kernel Regression
kr_result <- npreg(formula = Yes_Count ~ Age, bws = bw, data = alzDataAge)

# Extract Age column as a numeric vector
x_fit <- kr_result$eval[,1]  # If eval is a dataframe, extract first column
y_fit <- kr_result$mean      # Already a numeric vector
```

```{r,echo=FALSE,fig.align='center',fig.width=4.5,fig.height=3.5}
# Plot original data
plot(alzDataAge$Age, alzDataAge$Yes_Count, 
     main = "Kernel Regression with Optimal Bandwidth",
     xlab = "Age", ylab = "Positive Diagnosis Count", 
     pch = 16, cex=0.8)

# Add the fitted line
lines(x_fit, y_fit, col = "red", lwd = 2)
```

The Kernel regression follows a very similar trend to the B-spline model. It's slope between the 65 and 75 age group seems more smooth than those from the B-splines,which we're almost vertical, this smoothness is expected based on model assumptions. Higher in the data (after 75), the kernel model displays it's flexibility in fitting the data more accurately than the spline models.

Generalized cross validation is used to compare these three models due to the non-parametric nature of kernel regression. AIC is still calculated and compared for the two linear and polynomial spline models. No AIC is calculated for the kernel regression.

```{r GCVscore,echo=FALSE}
# Kernel Regression: Calculate residuals for Kernel Regression
residuals_kr <- alzData_countage$Yes_Count - kr_result$mean
RSS_kr <- sum(residuals_kr^2)  # Residual Sum of Squares
# We have the bandwidth from kr_result
bw_value <- kr_result$bw # Bandwidth value from your kernel regression result
# Estimate degrees of freedom for Kernel Regression using bandwidth and sample size
n_kr <- length(alzData_countage$Yes_Count)
# A simple way to approximate df for kernel regression using the bandwidth
df_kr <- n_kr * (bw_value^2) / 2  # This is a heuristic approximation
# Compute GCV for Kernel Regression
GCV_kr <- RSS_kr / (1 - df_kr / n_kr)^2
# For B-Spline (Degree 1)
residuals_bs1 <- residuals(bsplineProj)
RSS_bs1 <- sum(residuals_bs1^2)
df_bs1 <- sum(diag(bsplineProj$qr$qr^2))  # Degrees of freedom for B-Spline model (degree 1)
n_bs1 <- length(alzData_countage$Yes_Count)
# For B-Spline (Degree 3)
residuals_bs3 <- residuals(bsplineProjDeg3)
RSS_bs3 <- sum(residuals_bs3^2)
df_bs3 <- sum(diag(bsplineProjDeg3$qr$qr^2))  # Degrees of freedom for B-Spline model (degree 3)
n_bs3 <- length(alzData_countage$Yes_Count)
GCV_bs1 <- RSS_bs1 / (1 - df_bs1 / n_bs1)^2
GCV_bs3 <- RSS_bs3 / (1 - df_bs3 / n_bs3)^2
# Step 1: Extract AIC values for B-splines (already calculated)
AIC_bsdeg1 <- AIC(bsplineProj)
AIC_bsdeg3 <- AIC(bsplineProjDeg3)
# Now compare the GCV values
model_comparison <- data.frame(
  Model = c("Kernel Regression", "B-Spline (Degree 1)", "B-Spline (Degree 3)"),
  GCV = c(GCV_kr, GCV_bs1, GCV_bs3),
  AIC = c("", round(AIC_bsdeg1,3), round(AIC_bsdeg3,3))  # NA for Kernel Regression, since AIC is not applicable
)
kable(model_comparison, 
      col.names = c("Model", "Generalized Cross Validation (GCV)", "AIC"),
      caption = "Model Comparison: GCV and AIC",
      align = c("l", "r", "r"),
      digits = c(0, 4, 4))  # Adjust number of digits as needed
```

In summary, the table displays the third degree B-spline with the lowest AIC score but the highest GCV reflecting that it may be over fitting but accurately reflecting the relationship in the data. Since this relationship is a step function in relation to human age within \~15 year range, it makes sense to follow the lower AIC value and assume the B-spline model has the best performance in analyzing response within the range of the knots. The kernel model has the lower GCV reflecting it may be better at generalizing than other models but this claim is difficult to validate on ages outside the data's age range due to the nature of the data: it would be difficult to collect individuals over the age of \>100 years information and also less hard but still a task to compile and compare ages individuals of age \<50.

Overall all three models show a common structure of steps based on age range. Clearly age is significantly related to an Alzheimer's diagnosis and also clearly the diagnosis rate goes up with age in noticable jumps after the age of \~65.

## Quadratic Model Analysis of Body Mass Index and Alzheimer's

The initial modeling approach applied a linear regression model to predict Alzheimer's diagnosis. However, due to poor performance, logistic and quadratic models were introduced to better capture potential nonlinear relationships. Model selection was based on AIC and BIC, which indicated that the quadratic model outperformed the ordinary logistic model. Consequently, the quadratic model was selected as the foundation for subsequent analysis on BMI.

BMI was identified as a key variable for further modeling. This selection was based on two factors: (1) BMI is a well-distributed continuous variable in the data set, making it suitable for nonlinear exploration, and (2) previous research has suggested a correlation between BMI and cognitive function in the elderly, which may have physiological implications for Alzheimer's disease [@kang2021].

```{r BMI_diagnosis_plot, include=FALSE, cache=TRUE, fig.height=4,fig.width=9, fig.align='center'}
data <- alzDataEmily
data_count_bmi <- data %>%
  group_by(BMI) %>%
  summarise(
    Diagnosis_Yes_Count = sum(AlzheimerDiagnosis == "Yes", na.rm = TRUE),
    Total_Sample_Size = n()
  )
ggplot(data_count_bmi, aes(x = BMI)) +
  geom_bar(aes(y = Diagnosis_Yes_Count, fill = "Alzheimers Diagnosis = Yes"), stat = "identity", alpha = 0.6) +
  geom_bar(aes(y = Total_Sample_Size, fill = "Total Sample Size"), stat = "identity", alpha = 0.3) +
  scale_fill_manual(values = c("Alzheimers Diagnosis = Yes" = "blue", "Total Sample Size" = "red")) +
  labs(title ="Count of Alzhiemer's Diagnosis \n with Respect to BMI",
       x = "BMI",
       y = "Count",
       fill = "Legend") +
  theme_minimal()
```

```{r Add.Quad.terms, include=FALSE, cache=TRUE}
data$Age2 <- data$Age^2
data$BMI2 <- data$BMI^2
data$CognitiveTestScore2 <- data$CognitiveTestScore^2
data$EducationLevel2 <- data$EducationLevel^2
```

```{r Quad.Full.model, include=FALSE, cache=TRUE}
quadratic_full_model <- glm(AlzheimerDiagnosis ~ ., family = binomial, data = data)
summary(quadratic_full_model)
quad_full_aic <- AIC(quadratic_full_model)
```

```{r Quad.BMI, echo=FALSE, cache=TRUE, fig.align="center",fig.width=6,fig.height=3.5}
quad_bmi_model <- lm(Diagnosis_Yes_Count ~ BMI + I(BMI^2), data = data_count_bmi)
quad_pred <- predict(quad_bmi_model, newdata = data.frame(BMI = data_count_bmi$BMI), interval = "prediction")

plot(data_count_bmi$Diagnosis_Yes_Count ~ data_count_bmi$BMI,
     pch = 16,
     main = "Quadratic Model Diagnosis Count",
     xlab = "BMI",
     ylab = "Positive Diagnosis Count")
lines(data_count_bmi$BMI, quad_pred[, "fit"], col = "blue", lwd = 2)
lines(data_count_bmi$BMI, quad_pred[, "upr"], col = "red", lty = 2)
lines(data_count_bmi$BMI, quad_pred[, "lwr"], col = "red", lty = 2)

#summary(quad_bmi_model)
```

Visualization of BMI and Alzheimer's diagnosis revealed slight fluctuations in confirmed cases within the BMI range, particularly at slightly higher intermediate levels. The trend appeared nonlinear, though the fluctuations were not extreme. Based on this pattern, a quadratic model incorporating BMI and its squared term was constructed. The model results indicated a relatively flat parabolic fit, which better captured the local relationship between BMI and confirmed cases than the linear model. However, the quadratic model's flexibility was limited, potentially missing more complex nonlinear trends.

To enhance the model's adaptability, B-spline and Generalized Additive Model (GAM) techniques were introduced for further exploration.

```{r LM.BMI.Splines, include=FALSE, cache=TRUE}
bspline_deg1 <- lm(Diagnosis_Yes_Count ~ bs(BMI, knots = c(23, 28), degree = 1), data = data_count_bmi)
bs_pred1 <- predict(bspline_deg1, newdata = data.frame(BMI = data_count_bmi$BMI), interval = "prediction")
plot(data_count_bmi$Diagnosis_Yes_Count ~ data_count_bmi$BMI,
     pch = 16,
     main = "B-spline(degree = 1)",
     xlab = "BMI",
     ylab = "Diagnosis Count")
lines(data_count_bmi$BMI, bs_pred1[, "fit"], col = "blue", lwd = 2)
lines(data_count_bmi$BMI, bs_pred1[, "upr"], col = "red", lty = 2)
lines(data_count_bmi$BMI, bs_pred1[, "lwr"], col = "red", lty = 2)

bspline_deg2 <- lm(Diagnosis_Yes_Count ~ bs(BMI, knots = c(23, 28), degree = 2), data = data_count_bmi)
bs_pred2 <- predict(bspline_deg2, newdata = data.frame(BMI = data_count_bmi$BMI), interval = "prediction")
plot(data_count_bmi$Diagnosis_Yes_Count ~ data_count_bmi$BMI,
     pch = 16,
     main = "B-spline (degree = 2)",
     xlab = "BMI",
     ylab = "Diagnosis Count")
lines(data_count_bmi$BMI, bs_pred2[, "fit"], col = "blue", lwd = 2)
lines(data_count_bmi$BMI, bs_pred2[, "upr"], col = "red", lty = 2)
lines(data_count_bmi$BMI, bs_pred2[, "lwr"], col = "red", lty = 2)

bspline_deg3 <- lm(Diagnosis_Yes_Count ~ bs(BMI, knots = c(23, 28), degree = 3), data = data_count_bmi)
bs_pred3 <- predict(bspline_deg3, newdata = data.frame(BMI = data_count_bmi$BMI), interval = "prediction")
plot(data_count_bmi$Diagnosis_Yes_Count ~ data_count_bmi$BMI,
     pch = 16,
     main = "B-spline (degree = 3)",
     xlab = "BMI",
     ylab = "Diagnosis Count")
lines(data_count_bmi$BMI, bs_pred3[, "fit"], col = "blue", lwd = 2)
lines(data_count_bmi$BMI, bs_pred3[, "upr"], col = "red", lty = 2)
lines(data_count_bmi$BMI, bs_pred3[, "lwr"], col = "red", lty = 2)
```

Compared to the quadratic model, B-splines provide a more flexible function form by allowing smoothness and curvature control through the degree of the polynomial. The analysis considered degree values of 1 (piecewise linear), 2 (quadratic), and 3 (cubic). The results showed that: when **degree = 1**, the fit resembled segmented linear trends with a relatively stable overall pattern; when **degree = 2 or 3**, a slight upward trend was observed in the lower BMI range (18--22), suggesting some local nonlinear relationships. Although the overall trend was not dramatic, higher-order splines better captured small fluctuations in the data.

To further improve model flexibility and reduce reliance on manually selecting knots and degrees, GAM was employed. GAM allows automatic smoothing adjustments while preserving nonparametric flexibility. The analysis explored different smoothing levels by setting the number of basis functions (**k**) to 4, 6, and 10; when **k = 4 and k = 6**: The fitted curve's are nearly flat, indicating that lower degrees of freedom struggled to capture nonlinear characteristics, and when **k = 10**: The curve displayed more local fluctuations, with slight upward and downward trends across multiple BMI intervals, suggesting potential complex nonlinear structures.

```{r GamBMI4,echo=FALSE,cache=TRUE, fig.align='center',fig.width=6.5,fig.height=3.5}
par(mar = c(5, 6, 4, 8))  # Increase left margin and reduce right margin
# Fit the models
gam_k4 <- gam(Diagnosis_Yes_Count ~ s(BMI, bs = "cr", k = 4), data = data_count_bmi)
gam_pred4 <- predict(gam_k4, newdata = data.frame(BMI = data_count_bmi$BMI))

gam_k6 <- gam(Diagnosis_Yes_Count ~ s(BMI, bs = "cr", k = 6), data = data_count_bmi)
gam_pred6 <- predict(gam_k6, newdata = data.frame(BMI = data_count_bmi$BMI))

gam_k10 <- gam(Diagnosis_Yes_Count ~ s(BMI, bs = "cr", k = 10), data = data_count_bmi)
gam_pred10 <- predict(gam_k10, newdata = data.frame(BMI = data_count_bmi$BMI))

# Create the plot for all three models
plot(data_count_bmi$Diagnosis_Yes_Count ~ data_count_bmi$BMI,
     pch = 16, 
     main = "GAMs for Diagnosis Count by BMI",
     xlab = "BMI", 
     ylab = "Diagnosis Count", 
     col = rgb(0, 0, 0, 0.75),
     ylim = c(0, max(c(gam_pred4, gam_pred6, gam_pred10)) * 1.2))  # Adjust y-axis limit to include all lines

# Add the lines for each GAM
lines(data_count_bmi$BMI, gam_pred4, col = "red", lwd = 2)
lines(data_count_bmi$BMI, gam_pred6, col = "blue", lwd = 2)
lines(data_count_bmi$BMI, gam_pred10, col = "green", lwd = 2)

# Add a legend
legend("bottomright", 
       legend = c("GAM (k = 4)", "GAM (k = 6)", "GAM (k = 10)"), 
       col = c("red", "blue", "green"), 
       lwd = 2, 
       title = "GAM Models", 
       xpd = TRUE,  # Allow legend to be outside the plot
       inset = c(-0.45, 0))  # Adjust the legend position to be outside

par(mar = c(5, 4, 4, 2))
```

```{r CV_Gam, echo=FALSE, cache=TRUE}
set.seed(123)
cv_gam_k4 <- cv.glm(data_count_bmi, gam_k4, K = 10)
cv_gam_k6 <- cv.glm(data_count_bmi, gam_k6, K = 10)
cv_gam_k10 <- cv.glm(data_count_bmi, gam_k10, K = 10)
cv_results <- data.frame(
  Model = c("GAM (k=4)", "GAM (k=6)", "GAM (k=10)"),
  AIC = c(AIC(gam_k4), AIC(gam_k6), AIC(gam_k10)),
  CV = c(cv_gam_k4$delta[1], cv_gam_k6$delta[1], cv_gam_k10$delta[1])
)
kable(cv_results)
```

Although higher degrees of freedom improved visualization, they also increased the risk of over-fitting. Model selection was ultimately guided by AIC and cross-validation (CV) errors. Balancing AIC and CV, the GAM model with k = 6 demonstrated the best stability and predictive performance, making it the most suitable approach for modeling the relationship between BMI and Alzheimer's diagnosis.

According to the research results, the dynamic changes in BMI may reveal Alzheimer's risk better than the value at a single point in time [@kang2021].

In summary, the relationship between BMI and Alzheimer's risk may not only be nonlinear, but may also be affected by the weight change pattern in the time periods. In predictive modeling and risk intervention, BMI should be considered as an indicator together with age, genetic factors, cognitive ability and other variables to achieve more accurate early identification and prevention of Alzheimer's. However, results from the quadratic BMI variable in both GLM and GAM models did not find a significant relationship with Alzheimer's diagnosis. This suggests that while BMI-related features may provide useful insights, additional research is needed to determine their precise role in Alzheimer's risk assessment.

Finally, BMI solely as a variable to predict Alzheimer's diagnosis is not a significant factor.

## Age and Family History Interactions

```{r GLM_A.FH_int, include=FALSE, cache=TRUE}
set.seed(123)
model_af <- glm(AlzheimerDiagnosis ~ Age * FamilyHistory,data = data, family = binomial)
summary(model_af)
glm_af_aic<-model_af$aic
cv_model_af <- cv.glm(data, model_af, K = 10)
cv_model_af$delta
```

After evaluating the interaction of multiple variables, we found that the interaction term between Age and Family History was borderline significant in the Generalized Linear Model (p = 0.012). This suggests that the effect of age on Alzheimer's diagnosis may be influenced by family history. To explore this relationship further, we constructed a Generalized Additive Model (GAM) incorporating interaction terms for comparison.

```{r GAM_A.FH_int, include=FALSE, cache=TRUE}
set.seed(123)
gam_af <- gam(AlzheimerDiagnosis ~ s(Age, by = FamilyHistory, k = 5) + FamilyHistory, data = data, family = binomial)
plot(gam_af, pages = 1, se = TRUE)
gam_af_aic<-AIC(gam_af)
data$AlzheimerDiagnosis <- ifelse(data$AlzheimerDiagnosis == "Yes", 1, 0)
cv_gam_af <- cv.glm(data, gam_af, K = 10)
summary(gam_af)
#cv_gam_af$delta
```

```{r ModelCompare, echo=FALSE, cache=TRUE}
model_comparison <- data.frame(
  Model = c("GLM", "GAM"),
  AIC = c(AIC(model_af), AIC(gam_af)),
  CV = c(cv_model_af$delta[1], cv_gam_af$delta[1]))
kable(model_comparison)
```

In terms of model evaluation, the AIC and cross-validation error of the GAM model were better than those of the corresponding GLM model, indicating that GAM performed better in balancing fitting ability and generalization performance.

The results revealed that while the overall trends for individuals with and without family history were similar, the risk of Alzheimer's diagnosis increased rapidly after age 60 in both groups. However, individuals with a family history exhibited a higher risk between ages 50-60, and their overall risk curve was elevated. This indicates that family history significantly raises early risk levels. The individual regression coefficients of Age and Family History was significant (p \< 0.05) while the interaction term was negligible towards determining an Alzheimer's diagnosis, highlighting their strong predictive power in the model. Given their importance, we conducted further analysis on these variables without their interaction.

# Multivariate Model Analysis

Utilizing the general linear model regression from the [@MASS] package, a saturated logistic regression model was created to provide a benchmark for evaluating the fit for simpler models. The underlying objective is to find the most parsimonious model and identify the variables that explain an Alzheimer's Diagnosis.

As many of the variables were converted to factors as a result of them being categorical, the saturated model was 58 predictor variables along with the intercept term. Since there is such a large amount of predictor variables included, we tested for multicollinearity to see if any of the variables had a significant relationship to any other.

To detect multicollinearity between variables, we used Variance Inflation Factors (VIF) [@salmerón2018a] to quantify how much variance each of the regression coefficients is inflated by as a result of the linear regression model. Mathematically, the VIF for the $j$-th predictor variable ($X_j$) is calculated as:

$$
 \text{VIF}_j = \frac{1}{1 - R_j^2}
 $$

Here, $R_j^2$ represents the coefficient of determination obtained by regressing $X_j$ on all other predictor variables in the model. A higher VIF indicates greater multicollinearity, suggesting that the variable may be linearly dependent on other variables. A commonly agreed upon VIF that would indicate significant multicollinearity is 10. The result of this testing yielded no VIF value greater than 2 in the saturated model, and as such it is concluded that there exists little to no multicollinearity among all of the predictors in the data-set.

### Variable Selection

To find a more parsimonious model, three stepwise variable selection methods are performed -- each using the same function "step" from the MASS [@MASS] package. The stepwise algorithm iterates through the variables of the model passed into the function along with the direction and extracts and compares the Akaike's Information Criterion (AIC) for each variable to determine an optimal model fitted by maximum likelihood. This value is ultimately used to compare each model of the data set to analyze how well a model fits. For the purposes of finding a parsimonious model, we tested all three directions, namely: "both", "forward", and "backward", as we did not want to assume they would yield similar models. Results concluded that the three stepwise variable selection directions did result in the same model. Because of this, only the "forward" selection model will be referenced going forward for no other reason than it was the fastest in computation-time. Upon finding an optimal model, that AIC value is recorded in Table 5 (found in the conclusion).

```{r Full.Log.Reg, include=FALSE, cache=TRUE}
# Logistic Regression model on all variables

alz_logit_full <- glm(AlzheimerDiagnosis~ .
                  ,data = alzDataDylan
                  ,family = binomial)
sum(vif(alz_logit_full) >= 5) #Check if any vifs are >= 5.
#summary(alz_logit_full)
full_saturated_model.aic <- alz_logit_full$aic
# Country, Age, FamilyHistory, and GeneticRisk seem to be the only significant variables.
```

```{r Step.fwd, include=FALSE,cache=TRUE}
 
 # Intercept model
 alz_logit_int <- glm(AlzheimerDiagnosis~1,data=alzDataDylan,family="binomial")
  
 # Step-wise feature selection ("forward" direction)
 step.fwd.model<- step(alz_logit_int, direction = "forward", 
                       scope = list(lower = alz_logit_int, upper = alz_logit_full),
                       trace = F) 
 vif(step.fwd.model) # Check if any vifs exceed 5 in forward selection model. None exceed 2
 sum(vif(step.fwd.model) >= 5) # Check if any vifs exceed 5 in forward selection model. None exceed 2
 summary(step.fwd.model)
 step_fwd_model.aic <- step.fwd.model$aic
```

Once again, the VIF function is used to ensure that no multicollinearity was introduced with the reduction of predictor variables. There was no significant VIF values.

The GLM model as determined by the stepwise forward selection method suggests that variables: "Country", "Age", "FamilyHistory", "GeneticRisk", and "UrbanvsRuralLiving" variables provide the "best, optimal" model. The first four listed make some intuitive sense from our understandings of Alzheimer's Disease, however "UrbanvsRuralLiving" is somewhat strange intuitively. In addition, the algorithm that determined this reduced model included this variable even though it showed a borderline significant p-value of 0.0961 which is not less than 0.05 as is the common practice.

To further investigate this peculiarity, a statistical inference method called chi-squared testing found in the class textbook [@RegressionGLM] of section 8.3.1, was used on the deviance's of the models to assess if a variable can be removed without significantly affecting the model fit. To do this, the change in deviance measures the change between the log likelihood of the saturated model (in this case, the saturated model is the optimal model produced by the stepwise forward selection method) and the log likelihood of a reduced model. The reduced model is the same as the saturated model sans one variable. The deviance of each model was calculated for the purpose of showing the differences between a significant change vs. a non-significant change. If removing a variable results in a large increase in deviance (and a small p-value less than 0.05 to show significance), it suggests that the variable plays an important role in the model fit. If this is not the case, the variable can be removed without significantly affecting the model fit.

Results of deviance tests suggest that removing the Urban vs Rural Living variable from the model would not significantly affect the model fit supported by its p-value being 0.096, which is greater than 0.05.

Conclusions made from test: the best reduced model from feature selection methods could be further simplified without significantly affecting the model.

```{r Step.fwd.chisquared.tests, echo=FALSE,cache=TRUE,results='asis',eval=TRUE}
 ## Fit reduced models (each with one variable removed)
 
 no_country.model <- glm(AlzheimerDiagnosis ~ 
                           Age + FamilyHistory + GeneticRisk + UrbanvsRuralLiving, 
                         data = alzDataDylan, family = binomial(link = "logit"))
 
 no_age.model <- glm(AlzheimerDiagnosis ~ 
                       Country + FamilyHistory + GeneticRisk + UrbanvsRuralLiving,
                     data = alzDataDylan, family = binomial(link = "logit"))
 
 no_famHist.model <- glm(AlzheimerDiagnosis ~ 
                           Country + Age + GeneticRisk + UrbanvsRuralLiving, 
                         data = alzDataDylan, family = binomial(link = "logit"))
 
 no_genRisk.model <- glm(AlzheimerDiagnosis ~ 
                           Country + Age + FamilyHistory + UrbanvsRuralLiving, 
                         data = alzDataDylan, family = binomial(link = "logit"))
 
 no_urbanvsRural.model <- glm(AlzheimerDiagnosis ~ 
                           Country + Age + FamilyHistory + GeneticRisk, 
                         data = alzDataDylan, family = binomial(link = "logit"))
 
 
 ## Chi-Squared test: Compare the best feature selected reduced model to one-variable removed models
 ## If removing a variable results in a very large increase in deviance (and a very small p-value), it suggests that the variable plays an important role in the model fit.
sig.no_country <- pchisq(no_country.model$deviance-step.fwd.model$deviance,
                          df=no_country.model$df.residual-step.fwd.model$df.residual,
                          lower.tail = FALSE)
 
sig.no_age <- pchisq(no_age.model$deviance-step.fwd.model$deviance,
                      df=no_age.model$df.residual-step.fwd.model$df.residual,
                      lower.tail = FALSE)
 
sig.no_famHist <- pchisq(no_famHist.model$deviance-step.fwd.model$deviance,
                          df=no_famHist.model$df.residual-step.fwd.model$df.residual,
                          lower.tail = FALSE)
 
sig.no_genRisk <- pchisq(no_genRisk.model$deviance-step.fwd.model$deviance,
                          df=no_genRisk.model$df.residual-step.fwd.model$df.residual,
                          lower.tail = FALSE)
 
sig.no_urbanvsRural <- pchisq(no_urbanvsRural.model$deviance-step.fwd.model$deviance,
                          df=no_urbanvsRural.model$df.residual-step.fwd.model$df.residual,
                          lower.tail = FALSE)
 
 # Create a table of results
 chi.results <- data.frame(
   `Variable Removed` = 
     c("Country",
       "Age",
       "Family History",
       "Genetic Risk",
       "Urban vs Rural Living"),
   `Deviance Change` = 
     c(no_country.model$deviance-step.fwd.model$deviance
       ,no_age.model$deviance-step.fwd.model$deviance
       ,no_famHist.model$deviance-step.fwd.model$deviance
       ,no_genRisk.model$deviance-step.fwd.model$deviance
       ,no_urbanvsRural.model$deviance-step.fwd.model$deviance),
   `Degree Freedom Difference` = 
     c(no_country.model$df.residual-step.fwd.model$df.residual,
       no_age.model$df.residual-step.fwd.model$df.residual,
       no_famHist.model$df.residual-step.fwd.model$df.residual,
       no_genRisk.model$df.residual-step.fwd.model$df.residual,
       no_urbanvsRural.model$df.residual-step.fwd.model$df.residual),
   `P-value` = 
     c(sig.no_country,
       sig.no_age, 
       sig.no_famHist, 
       sig.no_genRisk, 
       sig.no_urbanvsRural)
 )
 
  kable(chi.results, 
       caption = "Chi-Squared Test Results" ,
       label = "log-odds-gam",
       col.names = c("Variable Removed","Deviance Change","DF Change","P-Value")) %>%
   kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                 full_width = FALSE) %>%
  column_spec(2, color = ifelse(chi.results$Deviance.Change < 10, "red", "black")) %>%  # Make "Deviance Change" column bold
  column_spec(4, color = ifelse(chi.results$P.value < 0.05, "black", "red"))  # Highlight significant P-values
```

The results of $\chi^2$ tests suggested that removing the "UrbanvsRuralLiving" variable from the model would not significantly affect the model fit supported by a small change in deviance and insignificant p-value = 0.096. This test motivated the removal of this variable from the optimal model produced by the stepwise algorithm, further simplifying the model and resulted in the inference that a persons geographical living demographic does not have a significant affect on an Alzheimer's diagnosis.

A QQ-plot of the residuals of the resulting reduced model allows for the assessment of whether the deviance residuals follow a standard normal distribution. We found that the residuals strongly deviated from the linear normal assumption at the tail-end of the model indicating that the model is not capturing some pattern(s) in the data.

```{r Reduced.model.check, echo=FALSE, cache=TRUE, fig.align='center', fig.height=3.25}
## Fit reduced model
alz_logit_reduced <- glm(AlzheimerDiagnosis~ 
                   Country+ Age+ FamilyHistory+ GeneticRisk
                  ,data = alzDataDylan
                  ,family = binomial)
#summary(alz_logit_reduced)
alz_logit_red.aic <- alz_logit_reduced$aic
#alz_logit_red.aic
plot(alz_logit_reduced, which = 2,
      main = "Residuals vs Fitted Values - Reduced Model", 
      col = "black", pch = 20, cex = 0.6, sub = "") # Customizing points with color and size
```

As there is some non-linear relationship present, a general additive model (GAM) is introduced to smooth this relationship to improve the models fit. Since "Age" is the only variable in the model that is not categorical, and from the results of the above analysis on age, we suspected that this variable's nonlinear relationship with Alzheimer's may be the reasons for this. This smoothing was performed by using cubic regression splines on "Age" instead of a simple linear term. The flexibility parameter "k" in the GAM model was adjusted and analyzed for values between 10-45.

```{r GAM.model, include=FALSE, cache=TRUE}
alz_logit_gam<- gam(AlzheimerDiagnosis~ 
                    Country+ FamilyHistory+ GeneticRisk+
                    s(Age, bs="cr",k=34)
                   ,data = alzDataDylan
                   ,family = binomial)
summary(alz_logit_gam)
alz_logit_gam.aic <- alz_logit_gam$aic
alz_logit_gam.aic
```

```{r Gam.plot, echo=FALSE, fig.height=3.5}
 plot(alz_logit_gam,
      main = "Estimated Smooth Effect of Age \n on the Log-Odds of Alzheimers Diagnosis"
      ,ylab="Estimated Log-Odds from Smooth Function")
```

The results of the Generalized Additive Model provides a scale estimate that assesses if over-fitting or under-fitting is present. An ideal value for this estimate is as close to 1 as possible. We found that a flexibility parameter of 34 achieved a scale estimate of 1, reducing the possibilities of over-fitting/under-fitting.

The results also highlight a non-linear relationship between age and Alzheimer's diagnosis. The smooth function applied to "Age" using cubic regression splines confirms this non-linearity, as seen in the estimated smooth effect plot. The sharp increase in log-odds around ages 66 and again around 76 suggests a critical transition period where the likelihood of an Alzheimer's diagnosis significantly increases around these ages. This pattern aligns with existing medical literature indicating that late-onset Alzheimer's risk increases exponentially with age.

The deviance residuals Q-Q plot shows that the residuals closely follow the theoretical quantiles, with minor deviations at the tails. This suggests that the model provides a reasonable fit to the data, though some small departures from normality may exist. The model summary further supports the importance of age, with the smooth term for "Age" having a highly significant p-value $(<2e-16).$

Beyond age, several categorical variables significantly contribute to the model. Family history and genetic risk are highly significant predictors, reinforcing their well-established roles in Alzheimer's risk.

```{r Log.odds, include=FALSE, cache=TRUE}
alz_gam.log_odds_ratio <- exp(coef(alz_logit_gam))
# Put into a table for clarity, sorted desc
 df_gam <- data.frame(Predictor = names(alz_gam.log_odds_ratio),
                      LogOdds_GAM = alz_gam.log_odds_ratio,
                      row.names = NULL)

# Replace spline component names for readability
df_gam$Predictor <- gsub("s\\(Age\\)\\.(\\d+)", "Age (Spline Component \\1)", df_gam$Predictor)

# Sort by the strongest log-odds in the reduced model
log_odds_table <- df_gam %>% arrange(desc(abs(LogOdds_GAM)))

## Put this table in the appendix
kable(log_odds_table, 
       caption = "Log-Odds of GAM Model Predictors",
       label = "log-odds-gam",
       col.names = c("Predictor", "Log Odds (GAM Model)")) %>%
   kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                 full_width = FALSE)
```

To further assess the impact of each predictor in the Generalized Additive Model (GAM), we calculated the log-odds ratios of all variables. These values indicate the relative change in the odds of an Alzheimer's diagnosis associated with each predictor. Recall that we found that a flexibility parameter of 34 provided the optimal conditions that maximized the log-odds. This means that the "Age" variable is split into 34 spline components and is shown in Table 6 in the appendix.

The strongest effects were observed in the spline components of age, particularly in the range of spline components 20--33, where the log-odds exceeded 3.9, with the highest being Age (Spline Component 22) at 4.99. This confirms that age remains the dominant risk factor, with the greatest effect occurring at older ages, aligning with known disease progression trends.

Among categorical predictors, Genetic Risk (log-odds = 3.77) and Family History (log-odds = 2.37) had the highest influence on Alzheimer's diagnosis, reinforcing the well-established role of genetic predisposition.

Country-level effects also revealed substantial variation in Alzheimer's diagnosis rates. Russia (log-odds = 1.82), India (log-odds = 1.78), and Brazil (log-odds = 1.65) showed much higher odds of diagnosis compared to the baseline country, while Canada (log-odds = 0.74), Norway (log-odds = 0.74), and Japan (log-odds = 0.69) exhibited lower odds. These disparities could stem from differences in healthcare access, diagnostic practices, or genetic and environmental factors.

Interestingly, spline components for lower ages (e.g., Spline Component 1-10, log-odds \< 0.35) had negligible effects, suggesting that younger individuals have minimal risk variations. The intercept term (log-odds = 0.34) represents the baseline probability of diagnosis before adjusting for predictors.

A complete table of log-odds ratios (Table 6) is provided in the appendix. These results underscore the importance of age, genetic risk, and geographic factors in predicting Alzheimer's, offering valuable insights for targeted risk assessment.

## Conclusion

This analysis sought to model the probability of an Alzheimer's diagnosis using a variety of demographic, behavioral, and genetic predictors, with a particular focus on the roles of age and family history. Through both generalized linear models (GLMs) and generalized additive models (GAMs), we aimed to identify the most parsimonious yet informative model to describe the risk factors associated with Alzheimer's disease.

Initial analyses suggested a borderline significant interaction between Age and Family History in the GLM, prompting further investigation using GAMs. The GAM confirmed that while the overall risk trends for individuals with and without a family history of Alzheimer's were similar, those with a family history exhibited an elevated risk between the ages of 50--60, with a sharp increase after age 60. These findings support the notion that a persons age and their family history significantly raises early risk levels, even though the interaction term itself did not contribute meaningfully to model performance.

A saturated logistic regression model was used as a baseline for multivariate analysis, with 58 predictors included. Multicollinearity was ruled out via VIF analysis, with all values below 2. To improve model parsimony, stepwise selection using AIC identified a reduced model including Country, Age, Family History, Genetic Risk, and Urban vs. Rural Living. However, further deviance testing showed that the Urban vs. Rural Living variable could be removed without a significant loss in model fit. This suggested that geographic living environment had minimal predictive power for Alzheimer's diagnosis in this data set.

The final reduced GLM showed signs of model misspecification, particularly in the tails of the deviance residuals. This indicated the presence of non-linear relationships, most notably involving Age. To address this, a GAM was fitted using cubic regression splines on Age. A flexibility parameter of k = 34 yielded an ideal scale estimate of 1, suggesting an optimal balance between over- and under-fitting.

The GAM revealed a clear non-linear effect of age, with sharp increases in log-odds of diagnosis occurring around ages 66 and 76 ---aligning with medical literature on late-onset Alzheimer's. Among categorical predictors, Genetic Risk (log-odds = 3.77) and Family History (log-odds = 2.37) showed the highest influence. Notably, country-specific effects revealed that individuals from Russia, India, and Brazil had considerably higher odds of diagnosis, potentially reflecting differences in healthcare access, genetic factors, or diagnostic practices.

```{r AIC.table,echo=FALSE,cache=TRUE}
 
 AIC.table <- data.frame(
   Model=c(
   "GLM - Full-Saturated Logistic Regression",
   "GLM - Quadratic Saturated",
   "GLM - Age * FamHist (Interaction)",
   "GAM - Age * FamHist (Interaction)",
   "GLM - Step Forward (Optimal)",
   "GLM - Reduced Optimal",
   "GAM - Reduced Optimal w/ Splines on Age"
   ),
   AIC=c(
     full_saturated_model.aic,
     quad_full_aic,
     glm_af_aic,
     gam_af_aic,
     step_fwd_model.aic,
     alz_logit_red.aic,
     alz_logit_gam.aic ),
   row.names = NULL
 )
 kable(AIC.table, 
       caption = "Model AIC Comparison",
       label = "aic-table",
       col.names = c("Model", "AIC")) %>%
   kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                 full_width = FALSE)
```

From the models and their respective AIC values that were created for this report, the one that captured the data and provided the strongest fit that lead to our conclusions was the GAM model that incorporated b-splines on "Age" because it had the lowest AIC value of 78140.18. Overall, the GAM improves model flexibility by capturing the non-linear age effect while still incorporating key categorical predictors. The adjusted R² value of 0.27 and deviance explained of 22.5% indicate a moderate fit, suggesting that while age and the included predictors are important, other factors that could be investigated are key genetic marker mutations in Alzheimer's patients, or other neurological-related variables might provide a stronger model for Alzheimer's detection.

The results of our analysis could benefit a range of stakeholders in both healthcare and public policy. Clinicians and medical researchers can use these insights to identify high-risk individuals---particularly those with a family history or genetic predisposition---and prioritize early screening or intervention strategies. Public health officials may also leverage these findings to tailor awareness campaigns and allocate resources more effectively in regions with elevated risk profiles. Additionally, the identification of non-linear age-related risk patterns can inform the development of age-targeted diagnostic tools and preventive programs, ultimately supporting efforts to delay the onset or progression of Alzheimer's disease.

\newpage

### Appendix

```{r echo=FALSE, cache=TRUE}
kable(log_odds_table, 
       caption = "Log-Odds of GAM Model Predictors",
       label = "log-odds-gam",
       col.names = c("Predictor", "Log Odds (GAM Model)")) %>%
   kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                 full_width = FALSE)
```

### Acknowledgements

We would like to express sincere gratitude to Dr. John R.J. Thompson for their guidance, support, and valuable feedback throughout the course of this project. Their expertise and encouragement were instrumental in shaping the direction of our analysis.

ChatGPT [@chatgpt], developed by OpenAI, was used in the preparation of this report to assist with drafting explanations, rephrasing text for clarity, and organizing sections. All analytical decisions, interpretations, and conclusions are solely our own.

\newpage

### Bibliography
